Estimating duration in the presence of misclassification

### 2024-02-09 transmission.tex
Updates from Overleaf

There are various advantages to using a mechanistic model to estimate transmission (see \cref{E-inc-prev:sec:infection-process}).
This section applies the mechanistic modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models} to estimate the transmission of SARS-CoV-2 in England.

The transmission part of the model follows \textcite{birrellRealtime}.
A novel approach (\cref{SEIR:sec:observation}) links the transmission model to prevalence data (here from the CIS, but the approach is generalisable).
In the mechanistic modelling literature, this part of the model is referred to as the observation model; however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.

This section uses the same notation as \cref{E-SEIR:sec:mechanistic-models}, following the infectious disease modelling literature rather than the statistical conventions used elsewhere in this thesis.

\subsection{Methods}

@@

\subsubsection{WAIFW matrix}

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocite{chenRole,vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

\todo[inline]{consider if this paragraph is too repetitive}

The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
A variety of different approaches can be taken to link a transmission model to observations.
In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively).
The method to do so is known as the observation model.
The formulation of this model is heavily dependent on the data available.
This means that there is a wide variety of models used.
This section  explains the novel modelling approach I propose for fitting to the CIS prevalence data.

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows a infected, pre-positive compartment, which I denote $p_0$, to be included simply.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).

@@

% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
An EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
It is shown schematically in \cref{SEIR:fig:EC}.
Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\omega_{X2})$.
% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
These distributions contain five parameters.
Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.

The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
Hence, the overall distribution is known as a Erlang-Coxian distribution.

>>>>

In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.

In common with \cref{E-SEIR:sec:mechanistic-models}, the notation in this section differs slightly from the rest of the thesis.

\subsection{Methods} \label{SEIR:sec:methods-application}

@@

\subsubsection{WAIFW matrix} \label{SEIR:sec:WAIFW-application}

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from infectious individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}(t)$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocite{chenRole,vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
%$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
This will link the model in the previous section to the CIS prevalence data.

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
This will lead to the model in FIGURE - with transition rates etc etc
The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).

@@

% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
An $l$-phase EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
It is shown schematically in \cref{SEIR:fig:EC}.
Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\omega_{X2})$.
% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
These distributions contain five parameters.
Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.

% The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
% Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
% The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
% Hence, the overall distribution is known as a Erlang-Coxian distribution.
